Future challenges for Research Software Engineering

### 2025-04-17 future_challenges_for_RSEng.tex
fix typo

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\texorpdfstring{\autref{1}}{},
Leyla Jael Castro\texorpdfstring{\autref{1}}{},
Gerasimos Chourdakis\texorpdfstring{\autref{1}}{},
Simon Christ\texorpdfstring{\autref{1}}{},
Jeremy Cohen\texorpdfstring{\autref{1}}{},
Jean-Noël Grad\texorpdfstring{\autref{1}}{},
Magnus Hagdorn\texorpdfstring{\autref{1}}{},
Toby Hodges\texorpdfstring{\autref{1}}{},
Jan Linxweiler\texorpdfstring{\autref{1}}{},
Frank Löffler\texorpdfstring{\autref{1}}{},
Michele Martone\texorpdfstring{\autref{1}}{},
Jan Philipp Thiele\texorpdfstring{\autref{1}}{},
Nick del Grosso\texorpdfstring{\autref{1}}{}
} % Authors and references to addresses
\institute{\autlabel{1} Fantasy University} % Institutes with labels
\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researchers day-to-day tasks, there are already people suggesting that RSE will
cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all domains.
Are we already approaching "the end of Research Software Engineering"?! While we believe that it is certainly too early to
already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

@@

% Main part of your article
% =========================
\section{Introduction}
This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
Nevertheless we hope, that we succeeded in writing up something that
\begin{itemize}
\item you like to read,
\item gives you something to think about,
\item motivates you to work on creating this better future.
\end{itemize}
In order to structure the discussion, we decided on a imaginable setting ten years from now. 
In this world we will consider the dimensions of 
\begin{itemize}
\item Evolution of the term RSEs
\item The development of academic institutions
\item What will the impact of AI be to the work of RSEs?
\item How will academic education have transformed by then
\item Ethics and social considerations
\end{itemize}
In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.

\section{The Situation}
In order to give a more explicit setting, imagine a world ten years from now.
Not too much, to be overwhelmingly far in the future, but distant enough for some changes to occur.
Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
But the world has moved on.
When they look back, what will they see as problems which are now solved?
Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.
\begin{story}{Introduction}{intro}
In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
of their future in respective small story boxes.
 \end{story}

\section{Aspects of this future}
It is difficult to structure the different aspects in describing the future. We decided to consider
\begin{itemize}
\item The evolution of the term RSE
\item How is education reshaped by the digital transformation
\item What will the impact of AI be to the work of RSEs?
\item Changes to the Academic Institutions
\item Ethics and social considerations
\item Impact of alternate career paths for RTPs
\end{itemize}

@@

Looking ahead over a 10 year timeframe, we hope that the RSE community will have
supported a number of changes and developments that help to make research software outputs
more sustainable, robust and maintainable. We see this happening in a number of ways. However,
we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.

With development of software tools or scripts becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

\begin{FramedParagraphWithFootnotes}
JN-NOTE:
In recent years, we saw the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible,
possibly through centrally-funded long-term positions at research institutions.
\end{FramedParagraphWithFootnotes}

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{acedemic}
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
\end{whatis}

\subsubsection{Which challenges will still persist}
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with succesful RSE Teams.
On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
The RSE is probably one of the first professionalizations in the academic midlevel that is able
to collectively voice their concerns. This enables the possibility to reshape universities.
If more people are employed at the midlevel, this makes a steadier pool of people
available that can integrate new people that are regularly coming into departments and know the local facilities.
On the other hand an upcoming challenge can be, how universities decide to design the contracts of people employed
in RSE units. will they be permanent, or will there also be an up-or-out mentality?

As outlined in [cite RSE group paper] not all universities need to opt for strengthening
the academic midlevel. Some might have chosen to set up external companies to outsource typical RSE tasks,
thereby providing career paths outside of academia.

FG-NOTE: By Nick. I've put it here, due to the realtion with the RTP question.

@@

The workshop participants found that in earlier learning stages,
details may be intentionally hidden,
fearing a diminishing capability to handle technical complexity in research environments.
As a result, RSEs invest more effort in making individual research software components
less complex with better user experience.
This made participants worry about growing mismatch of peoples' expectations.
Participants identified the over-reliance on generative AI tools as an extreme manifestation of this issue.
No concrete examples were given on which complexities are now hidden behind abstractions.

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible,
allowing the scientific community to address questions not previously in grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
but it is also important to learn how to abstract from details (create black boxes)
and build hierarchical systems from these abstractions, thereby managing complexity.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

\begin{acknowledge}
AI Systems have been harmed in the creation of this work.
We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!
\end{acknowledge}

>>>>

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\author{
Florian Goth\texorpdfstring{\autref{1}}{},
Leyla Jael Castro\texorpdfstring{\autref{1}}{},
Gerasimos Chourdakis\texorpdfstring{\autref{1}}{},
Simon Christ\texorpdfstring{\autref{1}}{},
Jeremy Cohen\texorpdfstring{\autref{1}}{},
Jean-Noël Grad\texorpdfstring{\autref{1}}{},
Magnus Hagdorn\texorpdfstring{\autref{1}}{},
Toby Hodges\texorpdfstring{\autref{1}}{},
Jan Linxweiler\texorpdfstring{\autref{1}}{},
Frank Löffler\texorpdfstring{\autref{1}}{},
Michele Martone\texorpdfstring{\autref{1}}{},
Jan Philipp Thiele\texorpdfstring{\autref{1}}{},
Nick del Grosso\texorpdfstring{\autref{1}}{}
} % Authors and references to addresses
\institute{\autlabel{1} Fantasy University} % Institutes with labels
\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all other domains.
+Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

@@

% Main part of your article
% =========================
\section{Introduction}
This opinion article stems from a Bof session at deRSE25 in Karlsruhe\cite{Goth2025EndRSEng}.
Therefore it stems from a discussion among the participants of this workshop and hence cannot claim that it represents the entire deRSE community.
Nevertheless we hope, that we succeeded in writing up something that
\begin{itemize}
\item is pleasant to read,
\item gives something to think about,
\item motivates engaging in work towards a better future.
\end{itemize}
In order to structure the discussion, we decided on an imaginable setting ten years from now.
In this world we will consider the dimensions of 
\begin{itemize}
\item Evolution of the term RSEs
\item The development of academic institutions
\item What will the impact of AI be to the work of RSEs?
\item How will academic education have transformed by then
\item Ethics and social considerations
\end{itemize}
In each aspect we will quickly summarise on what we want to have achieved by then, and what will be the challenges that arise from that.

\section{The Situation}
In order to give a more explicit setting, imagine a world ten years from now.
Not too much to be overwhelmingly far in the future, but distant enough for some changes to occur.
Over this timeframe Kim\cite{Anzt2021} and Kay\cite{Goth2024}
were able to ride the waves of their career in Research Software Engineering and are now pondering early retirement on a sunset blessed beach.
%MM-NOTE: the above scenario (early retirement) is contrary to all trends in retirement age development of public employees across the "global north".
But the world has moved on.
When they look back, what will they see as problems which are now solved?
Which tasks are persisting, where they are now happy that a younger generation is now taking up the baton and carries forward their work.
What issues have newly emerged, where they are now just happy to say: “Oh well… My successor takes care of that”.
\begin{story}{Introduction}{intro}
In order to bring this future to life these two personas will accompany us and we will shed light on certain aspects
of their future in respective small story boxes.
 \end{story}

\section{Aspects of this future}
It is difficult to structure the different aspects in describing the future. We decided to consider
\begin{itemize}
\item The evolution of the term RSE
\item How is education reshaped by the digital transformation
\item What will the impact of AI be to the work of RSEs?
\item Changes to the Academic Institutions
\item Ethics and social considerations
\item Impact of alternate career paths for RTPs (\emph{Research Technical Professional}s)
\end{itemize}

@@

Looking ahead over a 10-year timeframe, we hope that the RSE community will have
contributed to make research software and its outputs
more sustainable, robust and maintainable. We see this happening in a number of ways. However,
we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.

With development of software tools and scripted workflows becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
% MM-NOTE: if "this is vital" is referring to "software skills" it's not clear -- different paragraphs
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

@@

\begin{FramedParagraphWithFootnotes}
JN-NOTE:
In recent years, we saw the formation of RSE institutes and advocacy groups, such as
the UK Software Sustainability Institute\footnote{\url{https://www.software.ac.uk}},
the US Research Software Sustainability Institute\footnote{\url{https://urssi.us}},
the European Virtual Institute for Research Software Excellence\footnote{\url{https://everse.software}},
and the EOSC Opportunity Area Expert Group 7 on Research Software%
\footnote{\url{https://eosc.eu/opportunity-area-exp/oa7-research-software}}.
We expect these initiatives to help increase the visibility and accelerate
the professionalisation%
\footnote{MM: do we want to decide for UK or US English? Or keep it mixed? Not being a native English speaker I don't know how it reads.}
of the RSE role.
In addition, a multitude of RSE communities self-organised at the national
and regional level to support more viable RSE career paths in academia.
We can expect RSEs to become more easily accessible,
possibly through centrally-funded long-term positions at research institutions.
\end{FramedParagraphWithFootnotes}

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see, that researchers naturally accept them as part of their used services.
\end{whatis}

\subsubsection{Which challenges will still persist}
The workshop participants expect that there will still be budgeting and administrative hurdles to creating
permanent and well-funded RSE teams.
While we hope that the barrier to researchers diminishes, the workshop participants
still expect some work to do. We hope that the compartmentalization that occurs will help researchers reprioritize their workload.
An upcoming challenge will be that, with the emergence of more dedicated RSE units,
networking between them will come into the foreground, because there should be a strong need to share best practices
among institutions with successful RSE Teams.
On a more organisational level, we expect the sharing of models, to fit the different organisational constraints.

@@

\subsection{Impact of Alternate Career Paths for Researchers on the Structure of Research Institutes}
The RSE is probably one of the first professionalizations in the academic mid-level that is able
to collectively voice their concerns. This enables the possibility to reshape universities.
If more people are employed as mid-level, this makes a steadier pool of people
available that can integrate new people that are regularly coming into departments and know the local facilities.
On the other hand, an upcoming challenge how universities decide to design the contracts of people employed
in RSE units. Will they be permanent, or will there also be an up-or-out mentality?

As outlined in [cite RSE group paper] not all universities need to opt for strengthening
the academic mid-level. Some might have chosen to set up external companies to outsource typical RSE tasks,
thereby providing career paths outside of academia.

FG-NOTE: By Nick. I've put it here, due to the relation with the RTP question.

@@

The workshop participants found that in earlier learning stages,
details may be intentionally hidden,
fearing a diminishing capability to handle technical complexity in research environments.
As a result, RSEs invest more effort in making individual research software components
less complex with better user experience.
This made participants worry about growing mismatch of peoples' expectations.
Participants expected an over-reliance on generative AI tools as an extreme manifestation of this issue.
No concrete examples were given on which complexities are now hidden behind abstractions.

We acknowledge but would also like to challenge the observations of the participants.
We believe that ``we are standing on the shoulders of giants'':
By reducing the complexity of individual software components,
we are now able to develop systems of complexity and capability not previously possible,
allowing the scientific community to address questions not previously in grasp.
When it comes to teaching, it is true that students need to learn how to handle complexity,
but it is also important to learn how to abstract from details
by building hierarchical systems from these abstractions encapsulated in reusable components, thereby managing complexity.
It is clear that we need both RSEs that can dive deep into single projects of significant complexity,
as well as RSEs that can manage complex systems of almost black-box components,
and the future of RSE education should aim to prepare students for both roles.

@@

\begin{acknowledge}
% MM-NOTE: I'd keep the following sentence out, out of several reasons.
% AI Systems have been harmed in the creation of this work.
We thank all the participants of our workshop at deRSE25 in Karlsruhe for their valuable input!
\end{acknowledge}
